---
title: "Visualization and EDA"
author: "Linh Tran"
date: "10/5/2020"
output: github_document
---

```{r setup}
library(tidyverse)
library(ggridges)
```

```{r, warning = FALSE}
#Creating a df using meteo_pull_monitors in the `rnoaa` package

weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df
```

# Visualization with `ggplot2` - Part I

## Basic scatterplot
To create a basic scatterplot, need to map variables to the X and Y coordinate aesthetics, add geoms to define 

```{r, warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax)) + geom_point()

#New approach, same plot:

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()

#save the output to an object and modify/print later:

plot_weather = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) 

plot_weather + geom_point()
```
The basic scatterplot gave some useful information - the variables are related roughly as we'd expect, and there aren't any obvious outliers to investigate before moving on. 

## Advanced scatterplot - additional aesthetic mappings

`name` can be incorporated using the `color` aesthetic

```{r, warning = FALSE}

ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE)
  
#`name` can be incorporated using the `color` aesthetic

ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name))

#add a smooth curve and make the data points more transparent.
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) +
  geom_smooth(se = FALSE)
```
The curve gives a sense of the relationship between variables, and the transparency shows where data are overlapping. However, the smooth curve is for all the data but the colors are only for the scatterplot. That is because X and Y mappings apply to the whole graphic, but color is currently geom-specific. I am having a hard time seeing everything on one plot, so I'm going to add facet based on name as well.

```{r, warning = FALSE}
#add facet based on name

ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)


ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .2, size = 0.6) +      #alpha =.5 equals to 50% transparency
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)

ggplot(weather_df, aes(x = tmin, y = tmax,alpha = tmin, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

I've learned a lot about these data. However, the relationship between min and max temperature is now kinda boring, so I'd prefer something that shows the time of year. I also want to learn about precipitation.

```{r, warning = FALSE}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = 0.5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)

ggplot(weather_df, aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = 0.5) +
  geom_smooth(se = FALSE) + 
  facet_grid(name ~ .)
```

You can use a neat geom!

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_bin2d()

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_density2d() + 
  geom_point(alpha = .3)
```

#### Learning Assessment 1
Write a code chain focuses only on Central park, converts temperatures to Fahrenheit, make a scatterplot of min vs. max temperature, and overlay a linear regression line using options in geom_smooth()

```{r, warning = FALSE}
#this is my solution
library(dplyr)
ggplot(weather_df %>% 
  filter(name == "CentralPark_NY"), 
  aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name)) + 
  geom_smooth()

#this is the right solution
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_fahr = tmin * (9 / 5) + 32) %>% 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm", se = FALSE)
#Looks like there's a pretty linear relationship between min and max temperature in Central Park.
```

## Odds and Ends

```{r}
# There are lots of ways to mix and match elements, depending on your goals
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_smooth(se = FALSE) 

#When you're making scatterplot with lots of data, there's a limit to how much you can avoid overplotting using alpha levels and transparency. In these cases, `geom_hex()`, `geom_bin2d()`, or `geom_density2d()` can be handy
library(ggplot2)
ggplot(weather_df, aes(x = tmax, y = tmin)) +
  geom_hex()

#`color` worked for both geom_point() and geom_smooth() but `shape` only applies to points. 
```

#### Learning Assessment 2

```{r}
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin), color = "blue")
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin, color = "blue"))

#These lines don't produce the same result because in the 1st attempt, we're defining the color of the points by hand; in the 2nd attempt, we're implicitly creating a color variable that has the value "blue" everywhere; ggplot is then assigning colors according to this variable using the default color scheme.
```

## Univariate plots

Look at the distribution of single variables - this is an issue of learning some new geoms, and some new aesthetics. 

```{r, warning = FALSE}
ggplot(weather_df, aes(x = tmax)) + 
  geom_histogram()

#Can we add color
ggplot(weather_df, aes(x = tmax, color = name)) + 
  geom_histogram()

weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +
  geom_histogram(position = "dodge")

weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +
  geom_histogram() + 
  facet_grid(. ~ name)

#Play around with bin width and set the fill color using an aesthetic mapping. 

ggplot(weather_df, aes(x = tmax, fill = name)) +
  geom_histogram(position = "dodge", binwidth = 2)
 
#`position = "dodge` places the bars for each group side-by-side, but it can be hard to understand. prefer density plots over histogram. 

ggplot(weather_df, aes(x = tmax, fill = name)) +
  geom_density(alpha = 0.4, adjust = 0.5, color = "blue")
#`adjust` similar to `bindwidth` parameter. alpha = transparency level = 0.4 to make sure all densities appear. Lastly, adding `geom_rug()` can be helpful to show the raw data in addition to the density.

#boxplot
ggplot(weather_df, aes(x = name, y = tmax)) + geom_boxplot()

#violin plots
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), alpha = .5) + 
  stat_summary(fun = "mean", color = "blue", size = .2) +
  stat_summary(fun = "median", color = "red", size = .2)

#ridge plots - nice if you have lots of categories in which the shape of the distribution matters. 
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

#### Learning Assessment 3

```{r}
#Make plots that compare precipitation across location. Try histogram, density plot, boxplot, violin plot and ridgeplot

#density plot
ggplot(weather_df, aes(x = prcp)) + 
  geom_density(aes(fill = name), alpha = 0.5) 

ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_boxplot() 

#ridge plot
ggplot(weather_df, aes(x = prcp, y = name)) +
  geom_density_ridges(scale = 0.85)
  
#boxplot
ggplot(weather_df, aes(x = prcp, y = name)) +
  geom_boxplot()

# This is a tough variable to plot because of the highly skewed distribution in each location. Of these, I'd probably choose the boxplot because it shows the outliers most clearly. If the bulk of the data were interesting, I'd probably compliment this with a plot showing data for all precipiation less than 100, or for a data omitting days with no precipitation. 
weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, y = name)) +
  geom_density_ridges(scale = 0.85)
```

## Saving and embedding plots

```{r}
#Don't use export function - your figure will not be reproducible 
weather_plot = ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) 

ggsave("weather_plot.pdf", weather_plot, width = 8, height = 5)


#Embedding plots: The size of the figure created by R is controlled using 2 of the 3 chunk options `fig.width`, `fig.height`, `fig.asp`. Second is the size of the figure inserted into your document, which is controlled using `out.width` or `out.height`. 

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

#Embed at different size

```{r, fig.width = 6, fig.height = 3}
weather_plot
```

```{r, fig.width = 12, fig.height = 6}
weather_plot
```


# Visualization with `ggplot2` - Part II

```{r}
library(patchwork)
```

Revisit the scatterplot of tmax against tmin

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5)
```

#### Labels

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  )
```

#### Scales
Control over the location and specification of tick marks on the X or Y axis - You can use `scale_x_*` and `scale_y_*` where * depends on the type of variable (i.e. continuous vs discrete)

x and y scales 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ยบ C", "0", "15"))

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ยบC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")
```

Let's look at `scale_color_hue` 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_color_hue(name = "Location", h = c(100, 300))
```

#### Viridis package

```{r}
ggp_temp_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )

ggp_temp_plot
```

#### Themes

Shift the legend 

```{r}
ggp_temp_plot + 
  theme(legend.position = "bottom") 
```

Change the overall theme (background)

```{r}
ggp_temp_plot + theme_bw()

ggp_temp_plot + theme_minimal()

ggp_temp_plot + theme_classic()

ggp_temp_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### Setting options

These are options that I will put at the very beginning

```{r eval = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6  #height/width
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom")) #determine theme i want to exist everywhere

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_colour_continuous = scale_color_viridis_c
```


#### Data arguments in `geom` 

```{r}
central_park = 
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki = 
  weather_df %>% 
  filter(name == "Waikiki_HA")

ggplot(data = waikiki, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line()

ggplot(data = waikiki, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line(data = central_park)
```

#### `patchwork` 

Remember faceting?

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, fill = name)) +
  geom_density(alpha = .5) +
  facet_grid(.~name)
```

What happens when you want multipanel plots but can't facet..?

```{r}
tmax_tmin_plot =
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha =.5) +
  theme(legend.position = "none")

prcp_dens_plot =
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha =.5) +
  theme(legend.position = "none")

tmax_date_plot = 
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none")

tmax_tmin_plot + prcp_dens_plot 

tmax_tmin_plot + prcp_dens_plot + tmax_date_plot

(tmax_tmin_plot + prcp_dens_plot) / tmax_date_plot
```

#### Data manipulation

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha =.5)
```

name argument might not make perfect sense. It goes in alphabetical order according to the name. 
In the dataset I have, the name variable is character. When make plot, ggplot turns character variable to factor (1 = central park, 2 = waikiki, 3 = waterhole) and if I want to put them into different order, is there a way to overwrite that?

Control your factors

```{r}
weather_df %>% 
  mutate(
    name = factor(name),
    name = forcats::fct_relevel(name, c("Waikiki_HA"))
  ) %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha =.5)
```

An example reorder `name` according to `tmax` values in each name:

```{r}
weather_df %>%
  mutate(name = forcats::fct_reorder(name, tmax)) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

What if I wanted densities for tmin and tmax simultaneously? Three-panel plot showing densities for `tmax` and `tmin` within each location, facet panels across the `name` variable, and create separate densities for `tmax` and `tmin` in each panel. 

```{r}
weather_df %>% 
  select(name, tmax,tmin) %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observation",
    values_to = "temperature"
  ) %>% 
  ggplot(aes(x = temperature, fill = observation)) +
  geom_density(alpha =.5) +
  facet_grid(~name) +
  viridis::scale_fill_viridis(discrete = TRUE)
```

#### Revisiting PULSE data and create a plot showing BDI score across visits

```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  select(id, visit, everything()) %>%
  mutate(
    visit = recode(visit, "bl" = "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) %>%
  arrange(id, visit)

ggplot(pulse_data, aes(x = visit, y = bdi)) + 
  geom_boxplot()
```

#### Revisiting the pups

```{r}
pups_data = 
  read_csv("data/FAS_pups.csv") %>% 
  janitor::clean_names() %>% 
  mutate(sex = recode(sex, `1` = "male", `2` = "female"))

litters_data = 
  read_csv("data/FAS_litters.csv") %>% 
  janitor::clean_names() %>% 
  separate(group, into = c("dose", "day_of_tx"), sep = 3)

fas_data = left_join(pups_data, litters_data, by = "litter_number")

fas_data %>% 
  ggplot(aes(x = dose, y = pd_ears)) +
  geom_violin() +
  facet_grid(. ~ day_of_tx)

fas_data %>% 
  select(dose, day_of_tx, starts_with("pd_")) %>% 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome",
    values_to = "pn_day"
  ) %>% 
  drop_na() %>% 
  mutate(outcome = forcats::fct_relevel(outcome, "pd_ears", "pd_pivot", "pd_walk", "pd_eyes")) %>% 
  ggplot(aes(x = dose, y = pn_day)) +
  geom_violin() +
  facet_grid(day_of_tx ~ outcome)

fas_data %>% 
  select(sex, dose, day_of_tx, pd_ears:pd_walk) %>% 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome", 
    values_to = "pn_day") %>% 
  drop_na() %>% 
  mutate(outcome = forcats::fct_reorder(outcome, pn_day, median)) %>% 
  ggplot(aes(x = dose, y = pn_day)) + 
  geom_violin() + 
  facet_grid(day_of_tx ~ outcome)
```

# Exploratory analysis using data summaries

Load the weather data including month 

```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

#### group_by() makes grouping explicit and adds a layer to your data

```{r}
weather_df %>% 
  group_by(name)

weather_df %>% 
  group_by(name, month)

weather_df %>% 
  group_by(name, month) %>% 
  ungroup()
```

#### Counting things

Count month/ name observation

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n())

weather_df %>% 
  group_by(name, month) %>% 
  summarize(n_obs = n())
```

We can also use `count()` 

```{r}
weather_df %>% 
  count(month, name = "n_obs")

weather_df %>% 
  count(name, month, name = "n_obs")
```

**NEVER** use base R's `table`. 

```{r eval = FALSE}
weather_df %>% 
  pull(month) %>% 
  table()
```

Other helpful counters

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(),
    n_days = n_distinct(date)
  )
```

A digression on 2x2 table

```{r}
weather_df  %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE ~ ""
    )
  ) 
```

#### summarize() allows you to compute one-number summaries
